---
title: "Class05: Data Viz with GGPLOT"
author: "Li Ling (A15092789)"
format: gfm
---

Today we are playing with plotting graphics in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`,`hist()`, `boxplot()`, etc.)

There is also add-on packages, like **ggplot**,

```{r}
cars
```
```{r}
head(cars)
```

Let's plot this with "base" R:

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=12)
```

```{r}
hist(mtcars$mpg)
```
## GGPLOT

The main function in the ggplot2 package is `ggplot`. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`

> **N.B.** I never want to run `install.packages()` in my quarto source document!!!

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes(x=,y=)`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster

## Adding more layers

Let's add a line, a trend line given `geom_smooth()`, if you want a straight line, given `geom_smooth(method="lm)"`, then given `geom_smooth(method="lm", se=FALSE)` which hides the confidence interval shading, showing only the line, let's add subtitle and caption as well as custom axis labels.
```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE) +
  labs(
    title = "Silly Plot",
    x = "Speed (MPH)",
    y = "Distance (ft)",
    caption = "when is tea time anyways???") +
  theme_bw()
```

## Plot some expression data
Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset?

There are `r nrow(genes)` genes in this dataset.

> Q2. How many "up" regulated genes are there? A==B is A equal B, A!=B is A not equal B, A>B is A greater than B, A>=B is A greater than or equal to B, A<B is A less than B, A<=B is A less than or equal to B.

```{r}
#genes$State == "up" -->This will make the result a logical response
sum(genes$State == "up")
#This will give the number of "up" regulated genes
```
There are `r sum(genes$State == "up")` "up" regulated genes.

```{r}
table(genes$State)
```

> Q3. Make a scatter plot of `Condition1` vs `Condition2`, color the points by `State`, add a title and axis labels, make it a black and white theme.

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, color=Condition2) +
  geom_point()
 
```

```{r}
p <- ggplot(genes) +
  aes(Condition1, Condition2, color=State) +
  geom_point() +
  scale_color_manual(values=c("blue","grey","red")) 
 
```

Silly Example of adding labels
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, color=State, label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue","grey","red")) +
  geom_text()
 
```


```{r}
library(ggrepel)
ggplot(genes) +
  aes(Condition1, Condition2, color=State, label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue","grey","red")) +
  geom_text_repel(max.overlaps = 60)+
  theme_bw()
```

## Going further
Play with some different layers and the gapmider datasets...

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
#show the first 6 data points
head(gapminder)
#show the last 6 data points
tail(gapminder)
```


A first Plot
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point() +
  facet_wrap(~continent) +
  theme_bw() 
```

```{r}
p
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_violin()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_bin2d()
```


```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_density2d()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_smooth() +
  scale_x_log10() 
```







