---
title: "Class 10: Halloween Project"
author: "Li Ling (A15092789)"
format: pdf
---

## Importing Candy Data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

```{r}
flextable::flextable(head(candy))
```

```{r}
library(dplyr)

```





> Q1. How many different candy types are in this dataset?

There are 85 different candy types in this dataset.


```{r}
num_candy_types <- nrow(candy)
num_candy_types
```

> Q2. How many fruity candy types are in the dataset?
The functions dim(), nrow(), table() and sum() may be useful for answering the first 2 questions.

There are 38 fruity candy types in this dataset.

```{r}
num_fruity_candies <- sum(candy$fruity)
num_fruity_candies
```

```{r}
candy["Twix", ]$winpercent
```


## What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

The winpercent value for "Snickers" is 76.67%.

```{r}
candy["Snickers", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

The winpercent value for "Kit Kat" is 76.77%.

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The winpercent value for "Tootsie Roll Snack Bars" is 49.65%.

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

I think winpercent is different from other data becasue it is in percentage form while other columns are binary (0 or 1).

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents that the candy does not contain chocolate, while a one indicates that the candy contains chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Histogram of Winpercent Values", x = "Winpercent", y = "Frequency")
```


> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is slightly right-skewed (> 0 → right-skewed), indicating that there are more candies with lower winpercent values compared to those with higher winpercent values.

```{r}
# install.packages("moments")  # run this once if not installed
library(moments)

# Calculate skewness
skewness_value <- skewness(candy$winpercent, na.rm = TRUE)
skewness_value
```


> Q10. Is the center of the distribution above or below 50%?

Median is actually 47.83%, but Mean is 50.32%. For a skewed graph, we use median, so the center of distribution is below 50%.

```{r}
mean_winpercent <- mean(candy$winpercent, na.rm = TRUE)
mean_winpercent
```

```{r}
summary(candy$winpercent)
```



> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy has a higher average winpercent (60.92%) compared to fruity candy (44.12%).
```{r}
avg_chocolate_winpercent <- mean(candy$winpercent[candy$chocolate == 1], na.rm = TRUE)
avg_fruity_winpercent <- mean(candy$winpercent[candy$fruity == 1], na.rm = TRUE)
avg_chocolate_winpercent
avg_fruity_winpercent
```

```{r}
#Find all chocolate candy
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]
choc.candy

fruit.inds <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.inds,]
fruit.candy
#Extract their `winpercent` value
choc.win <- choc.candy$winpercent
fruit.win <- fruit.candy$winpercent
#Find the mean of these values
mean(choc.win)
mean(fruit.win)
```



> Q12. Is this difference statistically significant?

Yes, it is significant.

```{r}
choc <- candy$winpercent[candy$chocolate == 1]
fruit <- candy$winpercent[candy$fruity == 1]

t.test(choc, fruit, alternative = "two.sided")
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston baked Beans, Chiclets, Super Bubble, and Jawbusters.

```{r}
head(candy[order(candy$winpercent),], n=5)
```

```{r}
library(dplyr)

candy %>% arrange(winpercent) %>% head(5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat, and Snickers.
```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()

```
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()

my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?

Sixlets.

> Q18. What is the best ranked fruity candy?

Starburst.

## Taking a look at pricepercent


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip is the least popular one among the most expensive candies.

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar.

## Principal Component Analysis

The main function for PCA in R is `prcomp()`, and we want to `scale=TRUE`.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Let's look at our first main result figure - the "PC plot": PC1 vs PC2.

```{r}
ggplot(pca$x)+ aes(PC1, PC2, label=rownames(pca$x))+
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols)
```



```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
Don't forget about your variable "loadings" - hoe the original variable contributes to each PC.


```{r}
ggplot(pca$rotation)+ aes(PC1,rownames(pca$rotation))+
  geom_col()
```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus comes up very positive for PC1. It does make sense to me.































